A Fundamental Measure Theory Functional for Hard-Sphere Contact Densities

### 2012-07-06 papers/contact/paper.tex

@@

For the symmetric $S$ case:
\begin{align}
  g^S_\sigma(\rr) &= \frac{1}{n_0(\rr)^2}\int n^{(2)}(\rr - \rr', \rr
  + \rr')
  \frac{\delta(|\rr'| -\sigma/2)}{\pi\sigma^2}d\rr' \label{eq:gS} \\
  n_0(\rr) &= \int n(\rr')\frac{\delta(|\rr-\rr'|-\sigma/2)}{\pi\sigma^2} d\rr'
\end{align}
This functional gives a value averaged over spheres that touch at the
position $\rr$.  The density $n_0$ is one of the fundamental measures
of Fundamental Measure Theory (FMT), and seems ideal for treating
touching spheres, as illustrated in Figure~\ref{fig:n0}.

@@

And for the asymmetric $A$ case:
\begin{align}
  g^A_\sigma(\rr) &= \frac{1}{n(\rr)n_A(\rr)}
  \int n^{(2)}(\rr, \rr + \rr')
  \frac{\delta(|\rr'| - \sigma)}{4\pi\sigma^2}d\rr' \label{eq:gA} \\
  n_A(\rr) &= \int n(\rr')
  \frac{\delta(|\rr-\rr'|-\sigma)}{4\pi\sigma^2} d\rr' \label{eq:nA}
\end{align}
In this case, the two-particle density is averaged over spheres
touching a sphere that is aat the given position $\rr$.

@@

\section{Gross's asymmetrically averaged correlation functional}\label{sec:gross}
One existing approximation for the ccorrelation function is that of
Gross\cite{gross2009density}, which is of the asymmetrically averaged
variety ($g_\sigma^A$):
\begin{align}
  g_\sigma^\text{Gross,A}(\rr) &= \frac{1 - \frac{2\pi}{3}R^3n_A(\rr)}{\left(1 -
    \frac{4\pi}{3}R^3n_A(\rr)\right)^3}
\end{align}
where $n_A$ is the averaged density defined in Equation~\ref{eq:nA}.
This formula is arrived at by using the density averaged over all
spheres that could be touching a sphere at point $\rr$ in the
Carnahan-Starling equation for the correlation function at contact,
given in Equation~\ref{eq:cs-g}.

@@

\fixme{I (DJR) haven't yet edited this section!}
Yu and Wu developed a functional that should correspond to the contact
density at a given contact
position~$g^a$~\cite{yu2002fmt-dft-inhomogeneous-associating}.  We
will compare both their functional as well as ours with simulation
data.  Yu and Wu use a symmetric formulation for $g_\sigma$, but
instead of using $n_0$ as the associated density, it uses a density
given by
\begin{align}
  n_\text{Yu}(\rr) &= n_0(\rr) \zeta(\rr) \\
  \zeta &= 1 - \frac{\mathbf{n_2}\cdot\mathbf{n_2}}{n_2^2}
\end{align}
where the function $\zeta$ is a measure of local inhomogeneity at the
point of contact.  Given this definition of density, their $g_\sigma$ is
given by:
\begin{align}
  g_\sigma^\text{Yu} &= \frac{1}{1-n_3}
    + \frac32 \frac{R n_2}{3}\frac{\zeta}{(1-n_3)^2}
    + \frac12 \frac{\left(\frac{R n_2}{3}\right)^2 \zeta}{(1-n_3)^3}
    \\
  g_\sigma^\text{Yu} &= \frac{1}{1-n_3}
    + \frac12 \frac{R n_2\zeta}{(1-n_3)^2}
    + \frac1{18} \frac{R^2 n_2^2 \zeta}{(1-n_3)^3}
    \\
  g_\sigma^\text{Yu,S} &= \zeta^2\left(\frac{1}{1-n_3}
    + \frac12 \frac{R n_2\zeta}{(1-n_3)^2}
    + \frac1{18} \frac{R^2 n_2^2 \zeta}{(1-n_3)^3}\right)
\end{align}
I think the adjusted functional $g_\sigma^{\text{Yu,}S}$ can be directly
compared with our $g_\sigma^S$.  It is implemented in our code.

@@

\begin{figure}
  \includegraphics[height=10cm]{figs/walls-10}
  \caption{Density and contact densities at a wall with bulk filling
    fraction of 0.1.}
  \label{fig:walls-10}
\end{figure}

\begin{figure}
  \includegraphics[height=10cm]{figs/walls-30}
  \caption{Density and contact densities at a wall with bulk filling
    fraction of 0.3.}
  \label{fig:walls-30}
\end{figure}

\begin{figure}
  \includegraphics[height=10cm]{figs/walls-40}
  \caption{Density and contact densities at a wall with bulk filling
    fraction of 0.4.}
  \label{fig:walls-40}
\end{figure}

\begin{figure}
  \includegraphics[height=9cm]{figs/gnn-walls-10}
  \caption{Correlation multiplied twice by density with bulk filling
    fraction of 0.1.}
  \label{fig:gnn-walls-10}
\end{figure}

\begin{figure}
  \includegraphics[height=9cm]{figs/gnn-walls-30}
  \caption{Correlation multiplied twice by density with bulk filling
    fraction of 0.3.}
  \label{fig:gnn-walls-30}
\end{figure}

\begin{figure}
  \includegraphics[height=9cm]{figs/gnn-walls-40}
  \caption{Correlation multiplied twice by density with bulk filling
    fraction of 0.4.}
  \label{fig:gnn-walls-10}
\end{figure}

@@

\begin{figure}
  \includegraphics[height=5cm]{figs/contact-08-013}
  \caption{Density and contact densities of thirteen hard spheres in a
    spherical cavity with diameter 8. \sphereExplanation }
  \label{fig:sphere-8}
\end{figure}

\begin{figure}
  \includegraphics[height=5cm]{figs/contact-12-112}
  \caption{Density and contact densities of 112 hard spheres in a
    spherical cavity with diameter 12.  \sphereExplanation}
  \label{fig:sphere-12}
\end{figure}

\begin{figure}
  \includegraphics[height=5cm]{figs/contact-16-265}
  \caption{Density and contact densities of 265 hard spheres in a
    spherical cavity with diameter 16. \sphereExplanation}
  \label{fig:sphere-16}
\end{figure}

\begin{figure}
  \includegraphics[height=9cm]{figs/gnn-sphere-08-013}
  \caption{Correlation multiplied twice by density of 13 hard spheres in a
    spherical cavity with diameter 8.}
  \label{fig:gnn-sphere-08-013}
\end{figure}

\begin{figure}
  \includegraphics[height=9cm]{figs/gnn-sphere-12-112}
  \caption{Correlation multiplied twice by density of 112 hard spheres in a
    spherical cavity with diameter 12.}
  \label{fig:gnn-sphere-12-112}
\end{figure}

\begin{figure}
  \includegraphics[height=9cm]{figs/gnn-sphere-16-265}
  \caption{Correlation multiplied twice by density of 265 hard spheres in a
    spherical cavity with diameter 16.}
  \label{fig:gnn-sphere-16-265}
\end{figure}

@@

\begin{figure}
  \includegraphics[height=5cm]{figs/inner-4-30}
  \caption{Density and contact densities around a hard sphere with
    radius 1.}
  \label{fig:inner-4-30}
\end{figure}

\begin{figure}
  \includegraphics[height=5cm]{figs/inner-4-40}
  \caption{Density and contact densities around a hard sphere with
    radius 1.}
  \label{fig:inner-4-40}
\end{figure}

\begin{figure}
  \includegraphics[height=5cm]{figs/inner-8-30}
  \caption{Density and contact densities around a hard sphere with
    radius 3.}
  \label{fig:inner-8-30}
\end{figure}

@@

\begin{figure}
\includegraphics[width=5cm]{figs/gHS-vs-n}
\caption{A comparison of the various approximations to the contact
  density in the homogeneous limit.  As expected, they are all
  identical, and we won't include this plot in the paper (but it's
  here to verify that our code is working).}
\label{fig:gHS-vs-n}
\end{figure}

\begin{figure}
\includegraphics[width=5cm]{figs/free-energy}
\caption{A comparison of the various approximations to the free energy
  in the homogeneous limit.  Note that the White Bear functional uses
  a modified form of the Carnahan equation of state.  As expected,
  these are all identical, and we won't include this plot in the paper
  (but it's here to verify that our code is working).}
\label{fig:free-energy}
\end{figure}

>>>>

\fixme{Figure out whether to stick with both $R$ and $\sigma$, or move
to only $\sigma$ in the paper, which would have a certain appeal.}

@@

For the symmetric $S$ case:
\begin{align}
  g^S_\sigma(\rr) &= \frac{1}{n_0(\rr)^2}\int n^{(2)}(\rr - \rr', \rr
  + \rr')
  \frac{\delta(|\rr'| -\sigma/2)}{\pi\sigma^2}d\rr' \label{eq:gS} \\
  n_0(\rr) &= \int n(\rr')\frac{\delta(|\rr-\rr'|-\sigma/2)}{\pi\sigma^2} d\rr'
\end{align}
This functional gives a value averaged over spheres that touch at the
position $\rr$.  The density $n_0$ is one of the fundamental measures
of Fundamental Measure Theory (FMT), and is ideal for treating
touching spheres, as illustrated in Figure~\ref{fig:n0}.

@@

And for the asymmetric $A$ case:
\begin{align}
  g^A_\sigma(\rr) &= \frac{1}{n(\rr)n_A(\rr)}
  \int n^{(2)}(\rr, \rr + \rr')
  \frac{\delta(|\rr'| - \sigma)}{4\pi\sigma^2}d\rr' \label{eq:gA} \\
  n_A(\rr) &= \int n(\rr')
  \frac{\delta(|\rr-\rr'|-\sigma)}{4\pi\sigma^2} d\rr' \label{eq:nA}
\end{align}
In this case, the two-particle density is averaged over spheres
touching a sphere that is at the given position~$\rr$.

@@

\section{Gross's asymmetrically averaged correlation functional}\label{sec:gross}
One approximation for the correlation function is that of
Gross\cite{gross2009density}, which is of the asymmetrically averaged
variety ($g_\sigma^A$):
\begin{align}
  g_\sigma^\text{Gross,A}(\rr) &= \frac{1 - \frac{2\pi}{3}R^3n_A(\rr)}{\left(1 -
    \frac{4\pi}{3}R^3n_A(\rr)\right)^3}
\end{align}
where $n_A$ is the averaged density defined in Equation~\ref{eq:nA}.
This formula is arrived at by using the density averaged over all
spheres that could be touching a sphere at point $\rr$ in the
Carnahan-Starling equation for the correlation function at contact,
given in Equation~\ref{eq:cs-g}.

@@

Yu and Wu developed a functional for the correlation function
evaluated at contac, which should correspond to the symmetrically
averaged correlation
function~$g_\sigma^S$~\cite{yu2002fmt-dft-inhomogeneous-associating}.
Although Yu and Wu use a symmetric formulation for $g_\sigma$, instead
of using $n_0$ as the associated density, they use a density given by
\begin{align}
  n_\text{Yu}(\rr) &= n_0(\rr) \zeta(\rr) \\
  \zeta &= 1 - \frac{\mathbf{n_2}\cdot\mathbf{n_2}}{n_2^2}
\end{align}
where the function $\zeta$ is a measure of local inhomogeneity at the
point of contact.  Given this definition of density, their $g_\sigma$
is given by: \fixme{Which of the following two equations is more clear
  to you? We only want one.}
\begin{align}
  g_\sigma^\text{Yu} &= \frac{1}{1-n_3}
    + \frac32 \frac{R n_2}{3}\frac{\zeta}{(1-n_3)^2}
    + \frac12 \frac{\left(\frac{R n_2}{3}\right)^2 \zeta}{(1-n_3)^3}
    \\
  g_\sigma^\text{Yu} &= \frac{1}{1-n_3}
    + \frac12 \frac{R n_2\zeta}{(1-n_3)^2}
    + \frac1{18} \frac{R^2 n_2^2 \zeta}{(1-n_3)^3}
\end{align}
From this function, we extract a symmetric correlation function
\begin{align}
  g_\sigma^\text{Yu,S} &= \zeta^2\left(\frac{1}{1-n_3}
    + \frac12 \frac{R n_2\zeta}{(1-n_3)^2}
    + \frac1{18} \frac{R^2 n_2^2 \zeta}{(1-n_3)^3}\right)
\end{align}
which corresponds to our $g_\sigma^S$ and can be directly compared
with it.

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/walls-10}
  \caption{Density and contact densities at a wall with bulk filling
    fraction of 0.1.}
  \label{fig:walls-10}
\end{figure}

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/walls-30}
  \caption{Density and contact densities at a wall with bulk filling
    fraction of 0.3.}
  \label{fig:walls-30}
\end{figure}

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/walls-40}
  \caption{Density and contact densities at a wall with bulk filling
    fraction of 0.4.}
  \label{fig:walls-40}
\end{figure}

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/gnn-walls-10}
  \caption{Correlation multiplied twice by density with bulk filling
    fraction of 0.1.}
  \label{fig:gnn-walls-10}
\end{figure}

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/gnn-walls-30}
  \caption{Correlation multiplied twice by density with bulk filling
    fraction of 0.3.}
  \label{fig:gnn-walls-30}
\end{figure}

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/gnn-walls-40}
  \caption{Correlation multiplied twice by density with bulk filling
    fraction of 0.4.}
  \label{fig:gnn-walls-10}
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/contact-08-013}
  \caption{Density and contact densities of thirteen hard spheres in a
    spherical cavity with diameter 8. \sphereExplanation }
  \label{fig:sphere-8}
\end{figure}

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/contact-12-112}
  \caption{Density and contact densities of 112 hard spheres in a
    spherical cavity with diameter 12.  \sphereExplanation}
  \label{fig:sphere-12}
\end{figure}

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/contact-16-265}
  \caption{Density and contact densities of 265 hard spheres in a
    spherical cavity with diameter 16. \sphereExplanation}
  \label{fig:sphere-16}
\end{figure}

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/gnn-sphere-08-013}
  \caption{Correlation multiplied twice by density of 13 hard spheres in a
    spherical cavity with diameter 8.}
  \label{fig:gnn-sphere-08-013}
\end{figure}

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/gnn-sphere-12-112}
  \caption{Correlation multiplied twice by density of 112 hard spheres in a
    spherical cavity with diameter 12.}
  \label{fig:gnn-sphere-12-112}
\end{figure}

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/gnn-sphere-16-265}
  \caption{Correlation multiplied twice by density of 265 hard spheres in a
    spherical cavity with diameter 16.}
  \label{fig:gnn-sphere-16-265}
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/inner-4-30}
  \caption{Density and contact densities around a hard sphere with
    radius 1.}
  \label{fig:inner-4-30}
\end{figure}

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/inner-4-40}
  \caption{Density and contact densities around a hard sphere with
    radius 1.}
  \label{fig:inner-4-40}
\end{figure}

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/inner-8-30}
  \caption{Density and contact densities around a hard sphere with
    radius 3.}
  \label{fig:inner-8-30}
\end{figure}

@@

\begin{figure}
\includegraphics[width=\columnwidth]{figs/gHS-vs-n}
\caption{A comparison of the various approximations to the contact
  density in the homogeneous limit.  As expected, they are all
  identical, and we won't include this plot in the paper (but it's
  here to verify that our code is working).}
\label{fig:gHS-vs-n}
\end{figure}

\begin{figure}
\includegraphics[width=\columnwidth]{figs/free-energy}
\caption{A comparison of the various approximations to the free energy
  in the homogeneous limit.  Note that the White Bear functional uses
  a modified form of the Carnahan equation of state.  As expected,
  these are all identical, and we won't include this plot in the paper
  (but it's here to verify that our code is working).}
\label{fig:free-energy}
\end{figure}
